---
title: "Icefish"
output: rmarkdown::html_vignette
bibliography: Ref.bib
vignette: >
  %\VignetteIndexEntry{Icefish_2019}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width=6,
  fig.height=5
)
```

## Icefish Assessment

The Icefish assessment aims to determine the fishing mortality and hence a total allowable catch
that yields a prescribed two year escapement relative to an unfished population.

```{r,message=FALSE}
library(Grym)
```

## Data

Define the reference date that sets the start of the season
```{r}
SeasonDate <- as.Date("2015-12-01")
```

Define the survey date and the corresponding time increments, and the observed relative numbers and
biomass. Note that GYM averages over the start and end of the survey increment, so to match we
provide two survey increments.
```{r}
SurveyDate <- as.Date("2016-04-20")
surveyI <- as.numeric(SurveyDate-SeasonDate)+c(0,1)
surveyN <- c(29.87,90.53,1177.61,rep(0,7))
surveyB <- 3900.8173
```

Define the spawning date and the corresponding time increments
```{r}
SpawnDate <- as.Date("2016-11-30")
spawnI <- as.numeric(SpawnDate-SeasonDate) + c(0,1)
```


## Projection

The current strategy is to project forward two years for a range of potential fishing mortalities,
and then subsequently determine the mortality that yields the target relative escapement by inverse
interpolation.

The function `icefishPr` projects forward two years for a specified range of fishing mortalities.  A
projection is performed for the current year and scaled to the survey data to determine the initial
abundance for the following year. If the catch limit for the year has been reached the projection is
done assuming zero fishing mortality (as no further fishing can occur) but if some catch allocation
remains after the survey period, it is assumed the remaining allocation will be caught between the
end of the survey and the start of the following year.  The initial abundances are then projected
forward for two years for a range of fishing mortalities, and annual summaries collated.

The model assumes a von Bertalanffy relation between length and age, and a power law relation
between weight and length.  The user must supply that computes fishing selectivity as a function of
age.

The arguments are

* `F` - fishing mortalities to test
* `Catch` - remaining catch allocation after survey in survey year
* `surveyN` - relative numbers in each cohort from survey
* `surveyB` - biomass estimate from survey
* `surveyI` - increments (ie days of season) over which survey is taken
* `spawnI` - increments (ie days of season) over which spawning numbers and biomass are estimated
* `VB.t0`, `VB.K`, `VB.Linf` - parameters for von Bertalanffy length at age relation
* `WLa`, `WLb` - parameters for allometric weight at length relation
* `age.selectivity` - age selectivity function.
* `Fmax` - maximum allowable fishing mortality


```{r}
icefishPr <- function(M,F,Catch=0,surveyN,surveyB,surveyI,spawnI,
                      VB.t0=0.06671238,VB.K=0.36842178,VB.Linf=489.73706791,
                      WLa=9.157E-10,WLb=3.316,
                      age.selectivity=approxfun(c(0,2.5,3),c(0,0,1),rule=2),
                      Fmax=2.5) {
    
  ## Ensure 0 included in test fishing mortalities
  F <- sort(union(0,F))
  
  ## Two year projections of 10 age classes with a daily time step
  n.yr <- 2
  n.inc <- 365
  Ages <- 1:10
  Days <- seq(0,1,length=n.inc+1)
  
  ## Matrices of ages, lengths and weights for each day and age class
  as <- outer(Days,Ages,FUN="+")  
  ls <- vonBertalanffyAL(as,t0=VB.t0,K=VB.K,Linf=VB.Linf)
  ws <- powerLW(ls,a=WLa,b=WLb)

  ## Constant intra-annual natural mortality
  ms <- matrix(1,n.inc+1,length(Ages))
  Ms <- ctrapz(ms,1/n.inc)
  MMs <- M*Ms
  
  ## Within year fishing mortality is determined by an age based selectivity 
  fs <- array(age.selectivity(as),dim(as))
  Fs <- ctrapz(fs,1/n.inc)
  
  ### Projection to end of year from survey data
  if(Catch>0) {
    ## Adjust with-year fishing mortality for post-survey Catch
    fs0 <- rep.int(c(0,1),c(max(surveyI),n.inc+1-max(surveyI)))
    fs0 <- fs0/trapz(fs0,1/n.inc)*fs  
    Fs0 <- ctrapz(fs0,1/n.inc)
    pr0 <- projectC(ws,MMs,Fs0,fs0,Catch,surveyN,surveyI,surveyB,surveyI,yield=1,Fmax=Fmax)
    if(pr0$F==Fmax) warning("Target catch could not be recovered")
  } else {
    pr0 <- project(ws,MMs,0,0,surveyN,surveyI,surveyB,surveyI,yield=0)
    pr0$F <- 0
  }
  SSB0 <- meanStock(pr0$B,1,spawnI)

   
  ## Annual cohort totals
  d <- data.frame(Year=c(rep(0:n.yr,length(F))),F=0,Nf=0,Bf=0,Y=0,SSN=0,SSB=0,Escapement=0)
  k <- 0
  
  ## Project forward for prescribed fishing mortalities.
  for(Fk in F) {
    ## Reset to survey year
    pr <- pr0
    d[k <- k+1,] <- data.frame(Year=0,F=Fk,Nf=sum(final(pr$N)),Bf=sum(final(pr$B)),Y=sum(pr$Y),
                               SSN=meanStock(pr$N,1,spawnI),SSB=SSB0,Escapement=1)
    for(yr in seq_len(n.yr)) {
      ## Project
      N0 <- advance(pr$N)
      pr <- project(ws,MMs,Fk*Fs,Fk*fs,N0,yield=1)
      SSB <- meanStock(pr$B,1,spawnI)
      d[k <- k+1,] <- data.frame(Year=yr,F=Fk,Nf=sum(final(pr$N)),Bf=sum(final(pr$B)),Y=sum(pr$Y),
                                 SSN=meanStock(pr$N,1,spawnI),SSB=SSB,Escapement=SSB/SSB0)
    }
  }
  d
}
```




### No Remaining Allocation

Make projections for a range of fishing mortalities, assuming no catch
allocation remains after the survey 
```{r}
d <- icefishPr(M=0.4,F=seq(0.12,0.18,0.001),Catch=0,surveyN,surveyB,surveyI,spawnI)
head(d)
```

Compute the escapement and calculate the F the gives a two year
escapement of 0.75 relative to unfished
```{r,fig.show='hold'}
d2 <- d[d$Year==2,]
d2$RelEscape <- d2$Escapement/d2$Escapement[1]
F75 <- approx(d2$RelEscape[-1],d2$F[-1],0.75)$y
opar <- par(mar = c(5,4,4,4)+0.1)
plot(F~RelEscape,data=d2[-1,],type="l",col="grey60",
     xlab="Relative Escapement",ylab="Fishing Mortality")
abline(v=0.75,h=F75,col="dodgerblue")
points(F~RelEscape,data=d2[-1,],pch=16,col="grey60")
axis(3,at=0.75,col="dodgerblue")
axis(4,at=F75,labels=round(F75,4),col="dodgerblue",las=2)
par(opar)
```

Reproject for the target F
```{r}
d <- icefishPr(M=0.4,F=F75,Catch=0,surveyN,surveyB,surveyI,spawnI)
d
```


### Remaining Allocation

If there is an allocation of 200 tonnes remaining after the survey
```{r}
d <- icefishPr(M=0.4,F=seq(0.12,0.18,0.001),Catch=200,surveyN,surveyB,surveyI,spawnI)
head(d)
```

Compute the escapement and calculate the F the gives a two year escapement of 0.75 relative to unfished  
```{r,fig.show='hold'}
d2 <- d[d$Year==2,]
d2$RelEscape <- d2$Escapement/d2$Escapement[1]
F75 <- approx(d2$RelEscape[-1],d2$F[-1],0.75)$y
opar <- par(mar = c(5,4,4,5)+0.1)
plot(F~RelEscape,data=d2[-1,],type="l",col="grey60",
     xlab="Relative Escapement",ylab="Fishing Mortality")
abline(v=0.75,h=F75,col="dodgerblue")
points(F~RelEscape,data=d2[-1,],pch=16,col="grey60")
axis(3,at=0.75,col="dodgerblue")
axis(4,at=F75,labels=round(F75,5),col="dodgerblue",las=2)
par(opar)
```


Reproject for the target F
```{r}
d <- icefishPr(M=0.4,F=F75,Catch=200,surveyN,surveyB,surveyI,spawnI)
d
```

      
      

## Relative Escapement Targeting

The above process is easily extendend to automate the search for the fishing mortality that yields a
target relative escapement.

The function `icefishRE` uses `uniroot` to determine the fishing mortality that produces a desired
relative escapement and then projects forward for both zero fishing and the target fishing
mortality.  The arguments are

* `target` - target relative escapement
* `M` - natural mortality
* `F` - range of fishing mortalities to search
* `Catch` - remaining catch allocation after survey in survey year
* `surveyN` - relative numbers in each cohort from survey
* `surveyB` - biomass estimate from survey
* `surveyI` - increments (ie days of season) over which survey is taken
* `spawnI` - increments (ie days of season) over which spawning numbers and biomass are estimated
* `VB.t0`, `VB.K`, `VB.Linf` - parameters for von Bertalanffy length at age relation
* `WLa`, `WLb` - parameters for allometric weight at length relation
* `age.selectivity` - age selectivity function.
* `Fmax` - maximum allowable fishing mortality
* `tol` - error tolerance for `uniroot`


```{r}
icefishRE <- function(target,M,F,
                      Catch=0,surveyN,surveyB,surveyI,spawnI,
                      VB.t0=0.06671238,VB.K=0.36842178,VB.Linf=489.73706791,
                      WLa=9.157E-10,WLb=3.316,
                      age.selectivity=approxfun(c(0,2.5,3),c(0,0,1),rule=2),
                      Fmax=2.5,tol=1.0E-6) {

  ## Extract summary data from a projection
  annualSummary <- function(yr,F,pr) {
  }

  ## Ensure 0 included in test fishing mortalities
  F <- sort(union(0,F))
  
  ## Two year projections of 10 age classes with a daily time step
  n.yr <- 2
  n.inc <- 365
  Ages <- 1:10
  Days <- seq(0,1,length=n.inc+1)
  
  ## Matrices of ages, lengths and weights for each day and age class
  as <- outer(Days,Ages,FUN="+")  
  ls <- vonBertalanffyAL(as,t0=VB.t0,K=VB.K,Linf=VB.Linf)
  ws <- powerLW(ls,a=WLa,b=WLb)

  ## Constant intra-annual natural mortality
  ms <- matrix(1,n.inc+1,length(Ages))
  Ms <- ctrapz(ms,1/n.inc)
  MMs <- M*Ms
  
  ## Within year fishing mortality is determined by an age based selectivity 
  fs <- array(age.selectivity(as),dim(as))
  Fs <- ctrapz(fs,1/n.inc)
  
  ### Projection to end of year from survey data
  if(Catch>0) {
    ## Adjust with-year fishing mortality for post-survey Catch
    fs0 <- rep.int(c(0,1),c(max(surveyI),n.inc+1-max(surveyI)))
    fs0 <- fs0/trapz(fs0,1/n.inc)*fs  
    Fs0 <- ctrapz(fs0,1/n.inc)
    pr0 <- projectC(ws,MMs,Fs0,fs0,Catch,surveyN,surveyI,surveyB,surveyI,yield=1,Fmax=Fmax)
    if(pr0$F==Fmax) warning("Target catch could not be recovered")
  } else {
    pr0 <- project(ws,MMs,0,0,surveyN,surveyI,surveyB,surveyI,yield=0)
    pr0$F <- 0
  }
  ## Numbers at end of survey year - no recruitment
  N0Survey <- advance(pr0$N)
  SSB0 <- meanStock(pr0$B,1,spawnI)
  
  ## Project ahead and return final SSB 
  ProjectSSB <- function(F,target=0) {
    ## Project and compute SSB for final year
    N0 <- N0Survey
    for(yr in seq_len(n.yr)) {
      pr <- project(ws,MMs,F*Fs,F*fs,N0,yield=0)
      N0 <- advance(pr$N)
    }
    SSB <- meanStock(pr$B,1,spawnI)
    SSB-target
  }
  
  SSB1 <- ProjectSSB(0)
  r <- uniroot(ProjectSSB,F,target=target*SSB1)
  F <- c(0,r$root)
  
  ## Annual cohort totals
  d <- data.frame(Year=c(rep(0:n.yr,length(F))),F=0,Nf=0,Bf=0,Y=0,SSN=0,SSB=0,Escapement=0)
  k <- 0
  
  ## Project forward for prescribed fishing mortalities.
  for(Fk in F) {
    ## Reset to survey year
    pr <- pr0
    d[k <- k+1,] <- data.frame(Year=0,F=pr$F,Nf=sum(final(pr$N)),Bf=sum(final(pr$B)),Y=sum(pr$Y),
                               SSN=meanStock(pr$N,1,spawnI),SSB=SSB0,Escapement=1)
    for(yr in seq_len(n.yr)) {
      ## Project
      N0 <- advance(pr$N)
      pr <- project(ws,MMs,Fk*Fs,Fk*fs,N0,yield=1)
      SSB <- meanStock(pr$B,1,spawnI)
      d[k <- k+1,] <- data.frame(Year=yr,F=Fk,Nf=sum(final(pr$N)),Bf=sum(final(pr$B)),Y=sum(pr$Y),
                                 SSN=meanStock(pr$N,1,spawnI),SSB=SSB,Escapement=SSB/SSB0)
    }
  }
  d
}
```


### No Remaining Allocation

Estimate the fishing mortality that gives 75% escapement after two
years relative to unfished, assuming no catch allocation remains after
the survey
```{r}
d <- icefishRE(target=0.75,M=0.4,F=c(0,0.5),Catch=0,surveyN,surveyB,surveyI,spawnI)
d$RelEscapement <- d$SSB/d$SSB[1:3]
d
```

### Remaining Allocation

If there is an allocation of 200 tonnes remaining after the survey
```{r}
d <- icefishRE(target=0.75,M=0.4,F=c(0,0.5),Catch=200,surveyN,surveyB,surveyI,spawnI)
d$RelEscapement <- d$SSB/d$SSB[1:3]
d
```

