---
title: "Krill 2010"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Krill 2010}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6,
  fig.height=5
)
```

## Krill Assessment

This document aims to reproduce the Krill assessment process undertaken for WGEMM2010.


```{r,message=FALSE}
library(Grym)
library(ggplot2)
library(dplyr)
library(tidyr)
library(furrr)
set.seed(31)
```

## Recruitment model

Grym offers several variants of the proportional recruitment model with mutually interfaces -
select the analog of the model used in the GYM
```{r}
## Select recruitment model
prRecruitPars <- prRecruitParsGYM
prRecruitsQuantile <- prRecruitsQuantileGYM
prRecruits <- prRecruitsGYM
```

## Survey data


The primary inut data is the mean and variance of the estimated proportion of a "recruitment" age
class estimated from independent surveys to determine the properties of the recruitment model.
```{r}
R.mean <- 0.557
R.var <- 0.1^2
R.class <- 3
R.nsurveys <- 17
```


## Model

The `KrillProjection` function returns a function that generates a single set of random projections that
differ only in gamma.  The projections within each set use a common random selectivity and maturity
curves, virign biomass and recruitment series, so within a set the projections only differ by the
level of fishing applied.  The resulting function takes a single numerical argument that is used to
identify the set.

The arguments are 

* `R.mean`, `R.var` - the estimated mean and variance of the number of individuals in the reference
  age class as a proportion of all individuals of that age or older
* `R.class` - the reference age class.
* `R.nsurveys` - the number of surveys used to estimate `R.mean` and `R.var`.
* `gamma` - the gamma ratios to test
* `n.years` - the number of years to project ahead.

```{r}
KrillProjection <- function(R.mean,R.var,R.class,R.nsurveys,
                       gamma=c(0,0.04,0.08,0.1),n.years=20) {
  ## Daily time steps with 8 age classes
  nsteps <- 365
  Ages <- 0:7
  Days <- seq(0,1,length=nsteps+1)
  h <- 1/nsteps
  ## Spawning and monitoring interval
  spawnI <- 1:90
  monitorI <- 1:61
  ## Ages, length at age and weight at age
  ages <- outer(Days,Ages,FUN="+")
  ls <- vonBertalanffyRAL(ages,t0=0.0,K=0.45,Linf=60.8,f0=0,f1=91/365)
  ws <- powerLW(ls,1,3)
  ## Constant intra-annual natural mortality
  ms <- matrix(1,nsteps+1,length(Ages))
  Ms <- ctrapz(ms,h)
  Msf <- final(Ms)
  ## Within year fishing pattern - season is first 90 days
  fwy <- double(nsteps+1)
  fwy[1:90] <- 1
  fwy <- fwy/mean(fwy)
  ## Recruitment survey - parameters for the observed ratios
  ps0 <- prRecruitPars(Msf,R.mean,R.var,r=R.class)
  B0logsd <- 0.2
  
  ## This function performs the a projection for each prescibed gamma.
  function(run) {
    ## Length based maturity and selectivity - ramp width is constant
    ## but the midpoint is selected uniformly from a range.
    gs <- rampOgive(ls,runif(1,32,37),6)
    ss <- rampOgive(ls,runif(1,30,39),9)
    ## Construct fishing mortalities from season and selectivity
    fs <- fwy*ss
    Fs <- ctrapz(fs,h)
    Fsf <- final(Fs)
    ## Boostrap resample to account for uncertainty in R.mean and R.var
    repeat {
      Rbt <- prBootstrap(prRecruits,ps0,R.nsurveys,Msf,r=R.class)
      ps <- tryCatch(prRecruitPars(Msf,R.mean,R.var,r=R.class),error=function(e) e)
      if(!inherits(ps,"error")) break
    }
    ## Natural mortalities from proprtional recruitment model
    M <- ps$M
    MMs <- M*Ms
    ## Median spawning biomass estimated from 1000 samples
    R <- matrix(prRecruits(1000*length(Msf),ps),1000,length(Msf))
    ssb0 <- spawningB0S(R,gs,ws,Ms,M,spawn=spawnI)$median
    ## Stochastic initial age structure in the absence of fishing
    N0 <- ageStructureS(prRecruits(length(Msf),ps),Msf,M)
    ## Recruitment series
    Rs <- prRecruits(n.years,ps)
    ## Matrix of annual summaries
    n <- (1+n.years)*length(gamma)
    df <- matrix(0,n,11)
    colnames(df) <- c("Year","Gamma","R","N","B","B0","SSN","SSB","SSB0","Catch","F")
    ## Initial projection assuming no fishing
    pr0 <- project(ws,MMs,Nref=N0,yield=0)
    pr0$F <- pr0$Y <- 0
    ## Initial biomass in monitoring period + log Normal error
    b0 <- meanStock(pr0$B,period=monitorI)
    b0 <- rlnorm(1,log(b0)-B0logsd^2/2,B0logsd)
    
    k <- 0
    ## Project for each gamma ratio
    for(g in gamma) {
      ## Target catch
      catch <- g*b0
      ## Reset to virgin state
      pr <- pr0
      ssb <- spawningStock(pr$B,gs,spawnI)
      
      for(yr in 0:n.years) {
        if(yr > 0) {
          ## Recruitment depletion
          r <- min(1,ssb/(0.2*ssb0))
          ## Project over year
          N0 <- advance(pr$N,r*Rs[yr])
          pr <- projectC(ws,MMs,Fs,fs,catch,Nref=N0,yield=1,Fmax=1.5)
          #if(pr$F==1.5) return(NULL)
        }
        ssb <- spawningStock(pr$B,gs,spawnI)
        
        ## Collate annual summaries
        df[k<-k+1,] <- c(yr,g,initial(pr$N)[1],sum(initial(pr$N)),sum(initial(pr$B)),b0,
                         spawningStock(pr$N,gs,spawnI),ssb,ssb0,sum(pr$Y),pr$F)
      }
    }
    data.frame(Run=run,M=M,df)
  }
}
```



## Projection

Calling `KrillProjection` returns a function that generates a single "run" of the simulation
```{r}
sim <- KrillProjection(R.mean,R.var,R.class,R.nsurveys) 
```

The function takes a single argument that is used to label the results.  Within a single call to
`sim`, a common set of random maturity, selectivity, initial ages and recruitment series are used to
make projections for each requested gamma.
```{r}
df <- sim(1)
head(df)
tail(df)
```

For these parameters, the recruitment variability has greater impact
than increased fishing pressure
```{r}
library(ggplot2)
library(dplyr)
ggplot(df %>% mutate(Gamma=factor(Gamma)),aes(x=Year,y=N,colour=Gamma))+geom_line()
```

Generate 1000 runs and bind them into one large dataframe.  The runs are performed in parallel with
the `furrr` library
```{r}
plan(multiprocess)
system.time(df <- future_map_dfr(1:101,sim))
plan(sequential)
```

```{r}
df %>% group_by(Gamma,Run) %>% summarize(Dep=min(SSB/SSB0)) %>% summarize(Pr=mean(Dep < 0.2))
```
