---
title: "E. Superba"
output: rmarkdown::html_vignette
bibliography: Ref.bib
vignette: >
  %\VignetteIndexEntry{Krill_1996}
  %\VignetteEngine{rmarkdown::render}
  \usepackage[utf8]{inputenc}
---
```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width=6,
  fig.height=5
)

```

## E. superba

This document aims to reproduce the example presented in @GYM96.


```{r,message=FALSE}
library(Grym)
library(ggplot2)
library(dplyr)
library(tidyr)
library(furrr)
set.seed(31)
```


## Model

The `EsuperbaProjection` function returns a function that generates a single set of random projections that
differ only in gamma.  The projections within each set use a common random selectivity and maturity
curves, virign biomass and recruitment series, so within a set the projections only differ by the
level of fishing applied.  The resulting function takes a single numerical argument that is used to
identify the set.

```{r}
EsuperbaProjection <- function(gamma=c(0,0.1,0.136,0.15,0.2),n.years=20) {

  ## Daily time steps with 8 age classes
  nsteps <- 365   #Time steps
  Ages <- 0:7   #Age classes
  Days <- seq(0,1,length=nsteps+1)    #Cummulative time steps
  h <- 1/nsteps  #Proportion of the year for each time step. 

  ## Spawning and monitoring interval
  spawnI <- 121:213  #Time steps that spawning occurs on. 
  monitorI <- 92    #Time step of the monitoring interval. 

  ## Ages, length at age and weight at age
  ages <- outer(Days,Ages,FUN="+")  #Age at each time step in each age class. 
  ls <- vonBertalanffyRAL(ages,t0=0.0,K=0.45,Linf=60,f0=0,f1=93/365)   #Length at each time step in each age class. 
  ws <- powerLW(L=ls,a=1,b=3)   #Weight at each time step in each age class. 

  ## Constant intra-annual natural mortality
  ms <- matrix(data=1,nrow=nsteps+1,ncol=length(Ages))
  Ms <- ctrapz(fs=ms,h=h)
  Msf <- final(P=Ms)
  
  ## Within year fishing pattern - season is first 90 days
  fwy <- double(length=nsteps+1) #create vector of 0s the length of time steps
  fwy[31:120] <- 1 #Set the time steps of the fishing season to 1 
  fwy <- fwy/trapz(fs=fwy, h=h) #Scale the f in the fishing period to be 1 across the year

  B0logsd <- sqrt(log(1+0.3^2))
  
  ## This function performs the a projection for each prescibed gamma.
  function(run) {

    ## Length based maturity and selectivity - ramp width is constant
    ## but the midpoint is selected uniformly from a range.
    gs <- rampOgive(x=ls, x50=runif(n=1, min=34, max=40), xrange=12) #Maturity
    ss <- rampOgive(x=ls, x50=runif(n=1, min=38, max=42), xrange=10) #Selectivity

    ## Construct fishing mortalities from season and selectivity
    fs <- fwy*ss #Fishing pattern * selectivity
    Fs <- ctrapz(fs=fs, h=h) 
    Fsf <- final(P=Fs)

    ## Uniform natural mortalities 
    M <- runif(n=1,min=0.4,max=1.0) #Select mortality from within the range
    MMs <- M*Ms #Natural mortality by timestep

    ## Uniform (log) recruitment std dev. 
    sigmaR <- sqrt(log(1+runif(n=1, min=0.4, max=0.6)^2))

    ## Median spawning biomass estimated from 1001 samples
    R <- matrix(data=rlnorm(n=1001*length(Msf),meanlog = -sigmaR^2/2, sdlog=sigmaR),                    nrow = 1001,ncol = length(Msf)) #Recruitment matrix
    ssb0 <- spawningB0S(R,gs,ws,Ms,M,spawn=spawnI)$median #Median Spawning biomass

    ## Stochastic initial age structure in the absence of fishing
    N0 <- ageStructureS(R=rlnorm(n=length(Msf), meanlog = -sigmaR^2/2,
                                 sdlog = sigmaR), Msf=Msf, M=M)

    ## Recruitment series
    Rs <- rlnorm(n=n.years, meanlog = -sigmaR^2/2, sdlog = sigmaR)
    
    ## Matrix of annual summaries
    n <- (1+n.years)*length(gamma) #Sum number of years for each Gamma
    df <- matrix(data=0, nrow = n, ncol = 12+ncol(ages)) #create matrix of data
    colnames(df) <- c("Year","Gamma","R","N","B","B0","SSN","SSB","SSB0","SSB.Status","Catch","F",paste0("Age.Class.",1:ncol(ages))) #Set Column names
    
    ## Initial projection assuming no fishing
    pr0 <- project(ws=ws, MMs=MMs,Nref=N0,yield=0) 
    pr0$F <- pr0$Y <- 0
    
    ## Initial biomass in monitoring period + log Normal error
    b0 <- meanStock(P=pr0$B, period=monitorI)
    b0 <- rlnorm(n=1, meanlog = log(b0)-B0logsd^2/2, sdlog = B0logsd)
    
    k <- 0
    ## Project for each gamma ratio
    for(g in gamma) {
      ## Target catch
      catch <- g*b0

      ## Reset to virgin state
      pr <- pr0
      ssb <- spawningStock(P=pr$B, gs=gs, period=spawnI) 

      for(yr in 0:n.years) {
        
        if(yr > 0) {
          ## Recruitment depletion
          r <- min(1,ssb/(0.2*ssb0))
          ## Project over year
          N0 <- advance(N=pr$N,R=r*Rs[yr]) #roll age classes forward
          pr <- projectC(ws=ws,MMs=MMs,Fs=Fs,fs=fs,Catch=catch,Nref=N0,yield=1,Fmax=1.5) #Yearly projections for known catch.
          #if(pr$F==1.5) return(NULL)
        }
        ssb <- spawningStock(P=pr$B,gs=gs,period=spawnI)
      
        ## Collate annual summaries
        df[k<-k+1,] <- c(yr,g,initial(pr$N)[1],sum(initial(pr$N)),sum(initial(pr$B)),b0,
                        spawningStock(P=pr$N,gs=gs,period =spawnI) ,ssb, ssb0, ssb/ssb0 , sum(pr$Y) ,pr$F, initial(pr$N))
      }
    }
    data.frame(Run=run,M=M,df)
  }
}
```



## Projection

Calling `EsuperbaProjection` returns a function that generates a single "run" of the simulation
```{r}
sim <- EsuperbaProjection() 
```

The function takes a single argument that is used to label the results.  Within a single call to
`sim`, a common set of random maturity, selectivity, initial ages and recruitment series are used to
make projections for each requested gamma.
```{r}
df <- sim(1)
head(df)
tail(df)
```

For these parameters, the recruitment variability has greater impact
than increased fishing pressure
```{r}
library(ggplot2)
library(dplyr)
ggplot(df %>% mutate(Gamma=factor(Gamma)),aes(x=Year,y=N,colour=Gamma))+geom_line()
```

Generate 101 runs and bind them into one large dataframe.  The runs are performed in parallel with
the `furrr` library
```{r}
plan(multiprocess)
system.time(df <- future_map_dfr(1:101,sim))
plan(sequential)
#saveRDS(df, "Esup96_sim.rds")
```

## Results

```{r}
#df<-readRDS("Esup96_sim.rds")
```


```{r}
ggplot(df,aes(x=Year,y=N,group=Run))+
  geom_line(colour=rgb(0,0,0,0.02))+
  facet_wrap(~Gamma) + labs(y="Numbers")
```

```{r}
ggplot(df,aes(x=Year,y=B,group=Run))+
  geom_line(colour=rgb(0,0,0,0.02))+
  facet_wrap(~Gamma) + labs(y="Biomass")
```

```{r}
ggplot(df,aes(x=Year,y=SSN,group=Run))+
  geom_line(colour=rgb(0,0,0,0.02))+
  facet_wrap(~Gamma) + labs(y="Spawning Stock Numbers")
```

```{r}
ggplot(df,aes(x=Year,y=SSB,group=Run))+
  geom_line(colour=rgb(0,0,0,0.02))+
  facet_wrap(~Gamma) + labs(y="Spawning Stock Biomass")
```


```{r}
ssbstat<-df %>% group_by(Gamma, Year) %>% summarise(Med=median(SSB.Status), upr=quantile(SSB.Status, 0.975),lwr=quantile(SSB.Status, 0.025),q60=quantile(SSB.Status, 0.1),q40=quantile(SSB.Status, 0.9))

ggplot(df, aes(x=Year,y=SSB.Status,group=Run))+
  geom_line(colour=rgb(0,0,0,0.02))+
  geom_hline(yintercept=range(0.5,0.75), col="red", lty=3)+
  geom_ribbon(data=ssbstat, aes(x=Year, ymin=lwr, ymax=upr), inherit.aes = FALSE, alpha=0.2)+
  geom_ribbon(data=ssbstat, aes(x=Year, ymin=q40, ymax=q60), inherit.aes = FALSE, alpha=0.5)+
geom_line(data=ssbstat, aes(x=Year, y=Med), inherit.aes = FALSE, col="red")+
  facet_wrap(~Gamma)+
  theme_bw() + labs(y="Spawning Stock Status")

```


```{r}
names(df)
df1 <- df %>% filter(Year==0) %>% pivot_longer(cols=tidyselect::starts_with("Age.Class"),  names_prefix = "Age.Class.",names_to="Age",values_to="Na")
```



```{r}
d2 <- df1 %>% group_by(Gamma,Age) %>% summarize(N=mean(Na))
xtabs(round(N,4)~Gamma+Age,data=d2)
```



```{r}
df1 <- df %>% filter(Year==15) %>% pivot_longer(starts_with("Age.Class"),names_prefix = "Age.Class.",names_to="Age",values_to="Na")
```


```{r}
d2 <- df1 %>% group_by(Gamma,Age) %>% summarize(N=mean(Na))
xtabs(round(N,4)~Gamma+Age,data=d2)
```

We can use the dplyr function to calculate the depletion probability and the escapement for each Gamma tested. 

```{r}
df %>% group_by(Gamma,Run) %>% summarize(Dep=min(SSB/SSB0)) %>% summarize(Pr=mean(Dep < 0.2))

grymesc<- df %>%  group_by(Gamma) %>% filter(Year %in% max(Year)) %>% summarise(Med=median(SSB/SSB0))

grymtar<- grymesc %>% filter(Gamma==0) %>% mutate(Target=Med*0.75)

grymesc$Target<-grymtar$Target

grymesc

```


## References
